---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Read in the zip file
```{r readdata, echo=TRUE}
# Assume the data is in current working directory in a file named "activity.zip" 
activityData <- read.csv(unz("activity.zip", "activity.csv"))
```

Clean up and extend our data
```{r cleandata, echo=TRUE}
# the interval seems to be in 24 hour time 00:00 -> 23:55
# Add and extra column "time" that will convert date and interval into a
# POXIXct object for later data calulations
activityData$date <- as.Date(strptime(sprintf("%s %04d", 
                                      activityData$date, 
                                      activityData$interval), 
                              format = "%F %H%M", 
                              tz = "GMT")) 
```

## What is mean total number of steps taken per day?
```{r stepshistogram, echo=TRUE}
stepsPerDay<-tapply(activityData$steps,activityData$date,FUN=sum)
hist(stepsPerDay,breaks=20, main="Steps per Day",xlab="steps")
#require(ggplot2)
#qplot (data=stepsPerDay,geom="histogram")
```

```{r stepsmeanmedian, echo=TRUE}
# Remove the NAs while calculating the mean and median
stepsMM <- c(mean(stepsPerDay,na.rm=TRUE),median(stepsPerDay,na.rm=TRUE))
stepsMM <- setNames (stepsMM,c("Mean","Median"))
stepsMM
```


## What is the average daily activity pattern?

Average steps per interval
```{r averagedailyactivity, echo=TRUE}

# Calulate our averages over intervals ignoring NAs
avgDailyActivity <- data.frame(tapply( activityData$steps,
                                       activityData$interval,
                                       FUN=mean,
                                       na.rm=TRUE ))
avgDailyActivity$interval <-levels(as.factor( activityData$interval))
colnames(avgDailyActivity) <-c("Average.Steps","Interval")
with (avgDailyActivity, plot ( Average.Steps ~ Interval, 
                               type="l", 
                               main= "Average Daily Activity Pattern", 
                               xlab= "Interval",
                               ylab = "Average Steps per Interval")
)


```

Interval with the maximum number of steps (on average)
```{r maxavgsteps, echo=TRUE}
print (avgDailyActivity[which.max(avgDailyActivity$Average.Steps),],row.names=FALSE)
```

## Imputing missing values

Find the number of NAs
```{r countna, echo=TRUE}
sum(is.na(activityData$steps))
```

Imput the values for the NAs
```{r imputevalues, echo=TRUE}
# Copy the activity to a new data.frame object
# The NAs will be replaced with the mean value of the all the other 
# step measurements within the same interval
newActivityData <- activityData
newActivityData[is.na(newActivityData$steps),]$steps <- 
  round(tapply(X=newActivityData$steps,
               INDEX=newActivityData$interval,
               FUN=mean,
               na.rm=TRUE))
```

Create a histogram of the the new data calculate the mean and median
```{r newstepshistogram, echo=TRUE}
# sum up the total steps per day based on the the new data.frame 
# with the imputed steps per interval calaulated in the previous step
newStepsPerDay<-tapply(newActivityData$steps,newActivityData$date,FUN=sum)
hist(newStepsPerDay,breaks=20, main="Steps per Day",xlab="steps")
#require (ggplot2)
#ggplot(newStepsPerDay)

```

```{r newstepsmeanmedian, echo=TRUE}
# Calculate the mean and median (all NAs have been imputed)
newStepsMM <- c(mean(newStepsPerDay),median(newStepsPerDay))
newStepsMM <- setNames (newStepsMM,c("Mean","Median"))
newStepsMM
```
## Are there differences in activity patterns between weekdays and weekends?

Create a weekday/weekend factor
```{r daytype, echo=TRUE}
dow <- function (x) { 
  day<-weekdays(x)
  if (day=="Saturday"|| day=="Sunday") {
    return("weekend") 
  } else {
      return("weekday")
  }
}
activityData$daytype <- sapply(activityData$date,FUN=dow)
```

Average number of steps taken, averaged across all weekday days or weekend days
